### 2007-01-13 Documentation/Manual.tex
*** empty log message ***
JK - Some typos in Manual.tex corrected

\newcommand{\JHFSK}{{\sc JHF-SK}}
\newcommand{\NUMI }{{\sc NuMI}}
\newcommand{\ReactorI}{{\sc Reactor-I}}
\newcommand{\ReactorII}{{\sc Reactor-II}}
\newcommand{\JHFHK}{{\sc JHF-HK}}
\newcommand{\NuFactI}{{\sc NuFact-I}}
\newcommand{\NuFactII}{{\sc NuFact-II}}
\newcommand{\SPL}{{\sc SPL}}
\newcommand{\Beta}{{\sc $\beta$-Beam}}

@@

\vspace*{-4ex}
\GLOBES\ (``General Long Baseline Experiment Simulator'') is a flexible
software package to simulate neutrino oscillation 
long baseline and reactor experiments. On the
one hand, it contains a comprehensive abstract experiment definition
language (AEDL), which allows to describe most classes 
of long baseline experiments
at an abstract level. On the other hand, it provides a C-library to
process the experiment information in order to obtain oscillation
probabilities, rate vectors, and $\Delta \chi^2$-values. Currently, 
\GLOBES\ is available for GNU/Linux. Since the source code is included,
the port to other operating systems is in principle possible. The software
as well as up-to-date versions of this manual can be found at this URL:
 \verb+http://www.ph.tum.de/~globes+

\GLOBES\ allows to simulate experiments with stationary neutrino point sources, where each experiment is assumed to have only one neutrino source.
Such experiments are neutrino beam experiments and reactor experiments. 
Geometrical effects of a source distribution, such as in the sun or the 
atmosphere, can not be described. In addition, sources with a physically 
significant time dependencies  can not be studied, such as  supernov\ae. It 
is, however, possible to simulate beams with bunch structure, since the 
time dependence of the neutrino source is physically only important to suppress backgrounds.

@@

With the C-library, one can extract the $\Delta \chi^2$ for all defined 
oscillation channels for an experiment or any combination of experiments.
Of course, also low-level information, such as oscillation
probabilities or event rates, can be obtained. \GLOBES\ includes the
simulation of neutrino oscillations in matter with arbitrary matter 
density profiles, as well as it allows to simulate the matter density
uncertainty. As one of the most
advanced features of \GLOBES , it provides the technology to 
project the $\Delta \chi^2$, which is a function of all oscillation
parameters, onto any subspace of parameters by local minimization. 
This approach allows the inclusion of multi-parameter-correlations,
where external input (\eg, from solar parameters) can be imposed, too.
Applications of the projection mechanism include the projections onto the $\stheta$-axis and the $\stheta$-$\deltacp$-plane. In addition, all oscillation parameters can be kept free to precisely localize 
degenerate solutions.

@@

\GLOBES\ is developed for academic use. Thus, the \GLOBES\ Team would
appreciate being given academic credit for it. Whenever you use \GLOBES\
to produce a publication or a talk indicate that you have used \GLOBES\ and
please cite the following references~\cite{globes_paper,globes_paper_two}
\begin{quote}
P. Huber, M. Lindner and W. Winter\\
Simulation of long baseline neutrino oscillation experiments with \GLOBES\\
Comput. Phys. Commun. 167 (2005) 195, arXiv:\glbxxx.   
\end{quote}

\begin{quote}
P. Huber, J. Kopp, M. Lindner, M. Rolinec, and W. Winter\\
TO BE DEFINED, NEEDS TO BE UPDATED [WW]\\
arXiv:\glbxxxtwo.   
\end{quote}
but \emph{not} this manual. This manual itself is not a scientific 
publication and will not be submitted to a scientific journal. 
It will evolve during time since it is intended for 
regular revision. Besides that, many of the data which are used by \GLOBES\ 
and distributed together with it should be properly referenced. 
For details see below.

@@

It is recommended that you use the same style for your own input files, since, in case they are distributed, everybody will know how to correctly reference your work.

@@

As it is illustrated in \figu{GLOBES}, \GLOBES\ consists 
of several modules.
%
\begin{figure}[bht]
\begin{center}
\includegraphics[width=16cm]{GLOBES}
\end{center}
\caption{\label{fig:GLOBES} Different modules in \GLOBES .}
\end{figure}
%
\AEDL (``Abstract Experiment Definition Language'') is a language
to define experiments in form of ordinary text files. One or more of 
the resulting \AEDL\ files can then be processed together with supporting 
flux or cross section files by the user interface. The user interface
is a C-library, which loads one or more \AEDL\ file(s)
containing the experiment definition(s). The user interface is linked 
against the application software, and provides the user interface functions
for the intended experiment simulation.

The application 
software is, except from some example files, not part of \GLOBES , since
the evaluation of the experiment performance is often a matter of taste
and definition. In addition, the algorithms depend, especially for
high-precision instruments, very much on the oscillation parameters.
In general, it is quite simple to simulate superbeams and reactor
experiments. However, because of the more complicated topology, the
simulation of neutrino factories is much more difficult. In order
to demonstrate some of these difficulties, we present in this manual mainly
examples with neutrino factories. These examples can be found in
\Part~\ref{part:1} within the boxed pages. As complete files, they
are also available in the \GLOBES\ software package.

The \GLOBES\ software may have two target groups: 
Physicists, who are mainly interested in optimizing the potential
of specific experimental setups, and others, who are mainly
interested in the physics potential of different experiment types
from a theoretical point of view.
For the first group, \AEDL\ could be the most interesting aspect of
\GLOBES , where the user interface is only a tool to obtain specific
parameter sensitivities. In this case, \GLOBES\ could, serve as a
unified tool for the comparison and optimization of different experiment
 setups on equal footing, where
it is the primary objective to simulate the experiments as accurate
as possible. In addition, changes in experimental parameters, such as
efficiencies or the energy resolutions, can quickly be tested.
%
For the second user group, the pre-defined 
experiment definition files might already be sufficient to test
new conceptual approaches, and the user interface is the most interesting
aspect for sophisticated applications including correlations,
degeneracies, and multi-experiment setups. In either case, the \GLOBES\
software could serve as a platform for the exchange of experiment
definitions, and for an efficient splitting of work between
experimentalists and theorists.

@@

In \Part~\ref{part:2} of the manual, \AEDL\ is described. After an
introductory chapter, all functions are defined in greater detail.
This part might be more interesting for the experimental users who
want to modify or create \AEDL\ files. A useful tool in this context
is the software program \verb+globes+, which returns event rates and other
information for individual \AEDL\ files without further programming. 
For example, flux normalizations can with this tool be easily adjusted 
to reproduce the event rates of a specific experiment. It is described
in the last chapter of \Part~\ref{part:2}.

% THE FOLLOWING HAS CHANGED: C99, COMPILED WITH GCC; MAYBE ONLY PROBLEMS WITH C89
%
% \paragraph{Note:} All examples for application software in C do require
% a C++ compiler to be properly compiled. For pedagogical reasons, variable
% declarations are done at that place where the variable is needed for the
% first time, which is at variance with C syntax but not with C++ syntax.
% That is the only way in which the examples deviate form ISO C. Moreover
% the actual numerical values of the results of the examples may be different
% from the ones in this manual.

@@

\backmatter
\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

We would like to thank Martin Freund, who wrote the very first
version of a three-flavor matter profile treatment many years ago,
and Thomas Schwetz, who has been pushing the software to the edge in the 
past few years. Furthermore, we would like to thank Toshihiko Ota and WHO ELSE? [WW]
for using and testing unpublished new features of the software.
PH is especially thankful for the invaluable advice of Thomas Fischbacher on
many design issues in the early stage of the project.
% In addition, we would like to thank
% Mark Rolinec for his help to translate the experiment descriptions
% into \AEDL, and for creating the illustrations in this manuscript.
Finally, thanks to all the people who have been pushing this project
for many years, to the ones who have been continuing asking for the
publication of the software, and the referees of several of our
papers for suggestions which lead to improvements in the software.

>>>>

\newcommand{\JHFSK}{{\sc JHF-SK}}
\newcommand{\NUMI }{{\sc NuMI}}
\newcommand{\ReactorI}{{\sc Reactor-I}}
\newcommand{\ReactorII}{{\sc Reactor-II}}
\newcommand{\JHFHK}{{\sc JHF-HK}}
\newcommand{\NuFactI}{{\sc NuFact-I}}
\newcommand{\NuFactII}{{\sc NuFact-II}}
\newcommand{\SPL}{{\sc SPL}}
\newcommand{\Beta}{{\sc $\beta$-Beam}}
\newcommand{\NOVA}{{\sc NO$\nu$A}}
\newcommand{\TtoK}{{\sc T2K}}
\newcommand{\TtoHK}{{\sc T2HK}}
\newcommand{\DC}{{Double\sc Chooz}}

@@

\vspace*{-4ex}
\GLOBES\ (``General Long Baseline Experiment Simulator'') is a flexible
software package to simulate neutrino oscillation 
long baseline and reactor experiments. On the
one hand, it contains a comprehensive abstract experiment definition
language (AEDL), which allows to describe most classes 
of long baseline experiments
at an abstract level. On the other hand, it provides a C-library to
process the experiment information in order to obtain oscillation
probabilities, rate vectors, and $\Delta \chi^2$-values. Currently, 
\GLOBES\ is available for GNU/Linux. Since the source code is included,
the port to other operating systems is in principle possible. The software
as well as up-to-date versions of this manual can be found at this URL:
 \verb+http://www.mpi-hd.mpg.de/~globes+

\GLOBES\ allows to simulate experiments with stationary neutrino point sources, where each experiment is assumed to have only one neutrino source.
Such experiments are neutrino beam experiments and reactor experiments. 
Geometrical effects of a source distribution, such as in the sun or the 
atmosphere, can not be described. In addition, sources with a physically 
significant time dependence, such as supernov\ae, can not be studied. It 
is, however, possible to simulate beams with bunch structure, since the 
time dependence of the neutrino source is physically only important to suppress backgrounds.

@@

With the C-library, one can extract the $\Delta \chi^2$ for all defined 
oscillation channels for an experiment or any combination of experiments.
Of course, also low-level information, such as oscillation
probabilities or event rates, can be obtained. \GLOBES\ includes the
simulation of neutrino oscillations in matter with arbitrary matter 
density profiles, as well as it allows to simulate the matter density
uncertainty. As one of the most
advanced features of \GLOBES , it provides the technology to 
project the $\Delta \chi^2$, which is a function of all oscillation
parameters, onto any subspace of parameters by local minimization. 
This approach allows the inclusion of multi-parameter-correlations,
where external input (\eg, from solar experiments) can be imposed, too.
Applications of the projection mechanism include the projections onto the $\stheta$-axis and the $\stheta$-$\deltacp$-plane. In addition, all oscillation parameters can be kept free to precisely localize 
degenerate solutions.

@@

\GLOBES\ is developed for academic use. Thus, the \GLOBES\ Team would
appreciate being given academic credit for it. Whenever you use \GLOBES\
to produce a publication or a talk indicate that you have used \GLOBES\ and
please cite the following references~\cite{globes_paper,globes_paper_two}
\begin{quote}
P. Huber, M. Lindner and W. Winter\\
Simulation of long baseline neutrino oscillation experiments with \GLOBES\\
Comput. Phys. Commun. 167 (2005) 195, arXiv:\glbxxx,
\end{quote}

\begin{quote}
P. Huber, J. Kopp, M. Lindner, M. Rolinec, and W. Winter\\
TO BE DEFINED, NEEDS TO BE UPDATED [WW]\\
arXiv:\glbxxxtwo,
\end{quote}
but \emph{not} this manual. This manual itself is not a scientific 
publication and will not be submitted to a scientific journal. 
It will evolve during time since it is intended for 
regular revision. Besides that, many of the data which are used by \GLOBES\ 
and distributed together with it should be properly referenced. 
For details see below.

@@

It is recommended that you use the same style for your own input files, since then, in case they are distributed, everybody will know how to correctly reference your work.

@@

As it is illustrated in \figu{GLOBES}, \GLOBES\ consists 
of several modules.

\begin{figure}[bht]
\begin{center}
\includegraphics[width=16cm]{GLOBES}
\end{center}
\caption{\label{fig:GLOBES} Different modules in \GLOBES .}
\end{figure}
%
\AEDL (``Abstract Experiment Definition Language'') is a language
to define experiments in the form of ordinary text files. One or more of 
the resulting \AEDL\ files can then be processed together with supporting 
flux or cross section files by the user interface. The user interface
is a C-library, which loads one or more \AEDL\ file(s)
containing the experiment definition(s). The user interface is linked 
against the application software, and provides the user interface functions
for the intended experiment simulation.

The application 
software is, except for some example files, not part of \GLOBES , since
the evaluation of the experiment performance is often a matter of taste
and definition. In addition, the algorithms depend, especially for
high-precision instruments, very much on the oscillation parameters.
In general, it is quite simple to simulate superbeams and reactor
experiments. However, because of the more complicated topology, the
simulation of neutrino factories is much more difficult. In order
to demonstrate some of these difficulties, we present in this manual mainly
examples with neutrino factories. These examples can be found in
\Part~\ref{part:1} within the boxed pages. As complete files, they
are also available in the \GLOBES\ software package.

The \GLOBES\ software may have two target groups: 
Physicists, who are mainly interested in optimizing the potential
of specific experimental setups, and others, who are mainly
interested in the physics potential of different experiment types
from a theoretical point of view.
For the first group, \AEDL\ could be the most interesting aspect of
\GLOBES , where the user interface is only a tool to obtain specific
parameter sensitivities. In this case, \GLOBES\ could serve as a
unified tool for the comparison and optimization of different experiment
 setups on equal footing, where
it is the primary objective to simulate the experiments as accurate
as possible. In addition, changes in experimental parameters, such as
efficiencies or energy resolutions, can quickly be tested.
%
For the second user group, the pre-defined 
experiment definition files might already be sufficient to test
new conceptual approaches, and the user interface is the most interesting
aspect for sophisticated applications including correlations,
degeneracies, and multi-experiment setups. In either case, the \GLOBES\
software could serve as a platform for the exchange of experiment
definitions, and for an efficient splitting of work between
experimentalists and theorists.

@@

In \Part~\ref{part:2} of the manual, \AEDL\ is described. After an
introductory chapter, all functions are defined in greater detail.
This part might be more interesting for the experimental users who
want to modify or create \AEDL\ files. A useful tool in this context
is the executable program \verb+globes+, which returns event rates and other
information for individual \AEDL\ files without further programming. 
For example, flux normalizations can with this tool be easily adjusted 
to reproduce the event rates of a specific experiment. It is described
in the last chapter of \Part~\ref{part:2}.

% THE FOLLOWING HAS CHANGED: C99, COMPILED WITH GCC; MAYBE ONLY PROBLEMS WITH C89
%
% \paragraph{Note:} All examples for application software in C do require
% a C++ compiler to be properly compiled. For pedagogical reasons, variable
% declarations are done at that place where the variable is needed for the
% first time, which is at variance with C syntax but not with C++ syntax.
% That is the only way in which the examples deviate form ISO C. Moreover
% the actual numerical values of the results of the examples may be different
% from the ones in this manual.
%
% Comment by JK: A pedantic C89 compile will reject the example files (it would
% also reject GLoBES itself). With its default settings, gcc is relatively
% tolerant, so everything works with gcc. I assume that also the Intel
% compiler would be OK, but maybe Borland, Microsoft, or other compilers would
% produce problems.

@@

%\backmatter
\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

We would like to thank Martin Freund, who wrote the very first
version of a three-flavor matter profile treatment many years ago,
and Thomas Schwetz, who has been pushing the software to the edge in the 
past few years. Furthermore, we would like to thank Tommy Ohlsson,
Toshihiko Ota, and Julian Skrotzki for using and testing unpublished new features of the software.
PH is especially thankful for the invaluable advice of Thomas Fischbacher on
many design issues in the early stage of the project.
% In addition, we would like to thank
% Mark Rolinec for his help to translate the experiment descriptions
% into \AEDL, and for creating the illustrations in this manuscript.
Finally, thanks to all the people who have been pushing this project
for many years, to the ones who have been continuing asking for the
publication of the software, and the referees of several of our
papers for suggestions which lead to improvements in the software.
